Avoiding tipping points in the management of ecological systems: a non-parametric Bayesian approach

### 2013-06-20 inst/doc/nonparametric-bayes.Rmd
writing writing

```{r plot-options, echo=FALSE, cache=FALSE, include=FALSE}
options(xtable.print.comment=FALSE)
options(xtable.type = 'latex', table.placement="H")
opts_knit$set(progress = TRUE, verbose = TRUE)
opts_chunk$set(dev = c('pdf', 'png'), fig.width=5.5, fig.height=4,
 cache.path = 'cache/nonparametric-bayes/', cache=TRUE, external=TRUE)
opts_chunk$set(warning=FALSE, message=FALSE, comment=NA,
               tidy=FALSE)
toggle = "markup" # "hide" 
```

@@

Decision-theoretic methods often rely on simple parametric models of
ecological dynamics to compare the value of a potential sequence of
actions. Unfortunately, such simple models rarely capture the complexity
or uncertainty found in most real ecosystems. Non-parametric Bayesian
methods offer a promising statistical approach for predictive modeling
of ecological dynamics in regions of state space where the data is
adequate, while at the same time offering more flexible patterns with
greater uncertainty outside the observed data. This contrasts from simple
parametric models which provide relatively constant level of uncertainty
in regions with and without adequate data. The consequence of such
misplaced confidence outside the data can lead to highly undesirable
results that may be avoided with the more flexible non-parametric
Bayesian approach.

@@

Most management recommendations from the ecological literature are based
on (or at least motivated by) parametric models. Though in principle
these models reflect mechanistic underpinnings of biological interactions
involved, in practice real mechanisms are often unknown and the true
dynamics too complex to be captured by simple models. While
simple mechanistic models can nevertheless provide important insights
into possible dynamics -- for instance, demonstrating that a vaccine
does not have to be 100% effective to eliminate the transmission of a
virus  -- such approaches are not well-suited for use in
forecasting outcomes of potential management options.  Non-parametric 
approaches offer a more flexible alternative that can both more accurately
reflect the data available while also representing greater uncertainty
in areas (of state-space) where data is lacking.  Ecological research
and management strategy should pay closer attention to the opportunities 
and challenges nonparametric modeling can offer.

@@

<!-- cut?? -->

An active literature and growing computational power over the past several
decades have only marginally improved this situation.  Parametric or
structural uncertainty can be introduced only by increasing the state
space to reflect a distribution of possible parameters or
the degree of belief in each of a set of possible models Frequently,
the space of possible actions must then be reduced or the algorithms
adjusted by approximations to keep computations feasible.

<!-- cut?? -->

@@

```{r Figure1, dependson=c("assemble-models", "plot-options"), fig.cap="Graph of the inferred Gaussian process compared to the true process and maximum-likelihood estimated process.  Graph shows the expected value for the function $f$ under each model.  Two standard deviations from the estimated Gaussian process covariance with (light grey) and without (darker grey) measurement error are also shown.  The training data is also shown as black points.  (The GP is conditioned on 0,0, shown as a pseudo-data point). "}
```
```{r Figure1b, dependson=c("assemble-models", "par-fns", "plot-options")}
```

@@

```{r Figure2, dependson = c("assemble-opt", "plot-options"), fig.cap="The steady-state optimal policy (infinite boundary) calculated under each model.  Policies are shown in terms of target escapement, $S_t$, as under models such as this a constant escapement policy is expected to be optimal [@Reed1979]."}
```

@@

```{r Figure3, dependson=c("sim", "plot-options"), fig.cap="Gaussian process inference outperforms parametric estimates. Shown are 100 replicate simulations of the stock dynamics (eq 1) under the policies derived from each of the estimated models, as well as the policy based on the exact underlying model."}
```

@@

```{r Figure4, dependson=c("profits", "plot-options"), fig.cap="Histograms of the realized net present value of the fishery over a range of simulated data and resulting parameter estimates. For each data set, the three models are estimated as described above. Values plotted are the averages of a given policy over 100 replicate simulations. Details and code provided in the supplement."}
```

>>>>

```{r plot-options, cache=FALSE}
options(xtable.print.comment=FALSE)
options(xtable.type = 'latex', table.placement="H")
opts_knit$set(progress = TRUE, verbose = TRUE)
opts_chunk$set(dev = c('pdf', 'png'), fig.width=5.5, fig.height=4,
 cache.path = 'cache/nonparametric-bayes/', cache=TRUE, external=TRUE)
opts_chunk$set(warning=FALSE, message=FALSE, comment=NA,
               tidy=FALSE)
toggle = "markup" # "hide" 
theme_set(theme_bw(base_size=12))
```

@@

Decision-theoretic methods often rely on simple parametric models of
ecological dynamics to compare the value of a potential sequence of
actions. Unfortunately, such simple models rarely capture the complexity
or uncertainty found in most real ecosystems.

We demonstrate how nonparametric Bayesian models can provide robust,
nearly-optimal solutions to decision making under uncertainty when
_we don't know the correct model_ to use.

While methods that account
for _parametric_ uncertainty can be very successful with the right model,
structural uncertainty of not knowing what model best approximates the 
dynamics poses considerably greater difficulty.

Non-parametric Bayesian
methods offer a promising statistical approach for predictive modeling
of ecological dynamics in regions of state space where the data is
adequate, while at the same time offering more flexible patterns with
greater uncertainty outside the observed data. This contrasts from simple
parametric models which provide relatively constant level of uncertainty
in regions with and without adequate data. The consequence of such
misplaced confidence outside the data can lead to highly undesirable
results that may be avoided with the more flexible non-parametric
Bayesian approach.

@@

Decision making under uncertainty is a ubiquitous challenge of natural 
resource management and conservation.

<!-- Models for decision-making -->
Decision-theoretic tools require a model that can assign probabilites of
future states (e.g. stock size of a fishery) given the current state and a
proposed action (e.g. fishing harvest or effort).  The decision maker then
seeks to determing the course of actions (also refered to as the policy)
that maximizes the expected value of some objective function (such as
net present value derived from the resource over time.  (The approach
can be adapted to permit alternatives to maximizing expectation, such
as minimizing the maximum cost or damage that might be incurred. See
@Polasky2011).
Management frequently faces a sequential decision-making problem -- after 
selecting an action, the decision-maker may recieve new information about 
the current state and must again choose an appropriate action -- i.e. setting
the harvest limits each year based on stock assessments the year prior.

<!-- Parametric models -->

<!-- Parametric uncertainty -->

<!-- Comparing models -->
The nature of decision-making problems provides a convenient way to compare 
models.  Rather than compare models in terms of best fit to data or fret over
the appropriate penalty for model complexity, model performance is defined 
in the concrete terms of the decision-maker's objective function, which we
will take as given. (Much argument can be made over the 'correct' objective
function, e.g. how to account for the social value of fish left in the sea
vs. the commercial value of fish harvested; see @Halpern2013 for further 
discussion of this issue.  Alternatively, we can always compare model performance
across multiple potential objective functions.)  The decision-maker
does not necessarily need a model that provides the best mechanistic understanding
or the best long-term outcome, but rather the one that best estimates the 
probabilites of being in different states as a result of the possible actions.

<!-- Structural Uncertainty -->

 While
simple mechanistic models can nevertheless provide important insights
into long-term outcomes, such approaches are not well-suited for use in
forecasting outcomes of potential management options.  Non-parametric 
approaches offer a more flexible alternative that can both more accurately
reflect the data available while also representing greater uncertainty
in areas (of state-space) where data is lacking.

We demonstrate how
a Gaussian Process model of stock recruitment can lead to nearly optimal
management through stochastic dynamic programming, comperable to knowing
the correct structural equation for the underlying simulation.  Meanwhile,
parametric models that do not match the underlying dynamics can perform 
very poorly, even though they fit the data as well as the true model.

Ecological research
and management strategy should pay closer attention to the opportunities 
and challenges nonparametric modeling can offer.

@@

@@

```{r Figure1, dependson=c("assemble-models", "plot-options"), fig.cap="Graph of the inferred Gaussian process compared to the true process and maximum-likelihood estimated process.  Graph shows the expected value for the function $f$ under each model.  Two standard deviations from the estimated Gaussian process covariance with (light grey) and without (darker grey) measurement error are also shown.  The training data is also shown as black points.  (The GP is conditioned on 0,0, shown as a pseudo-data point). ", include=FALSE}
```
```{r Figureb, dependson=c("assemble-models", "par-fns", "plot-options"), toggle = 1}
```

@@

```{r Figure2, dependson = c("assemble-opt", "plot-options"), fig.cap="The steady-state optimal policy (infinite boundary) calculated under each model.  Policies are shown in terms of target escapement, $S_t$, as under models such as this a constant escapement policy is expected to be optimal [@Reed1979].", toggle = 1}
```

@@

```{r Figure3, dependson=c("sim", "plot-options"), fig.cap="Gaussian process inference outperforms parametric estimates. Shown are 100 replicate simulations of the stock dynamics (eq 1) under the policies derived from each of the estimated models, as well as the policy based on the exact underlying model.", toggle = 1}
```

@@

```{r Figure4, dependson=c("profits", "plot-options"), fig.cap="Histograms of the realized net present value of the fishery over a range of simulated data and resulting parameter estimates. For each data set, the three models are estimated as described above. Values plotted are the averages of a given policy over 100 replicate simulations. Details and code provided in the supplement.", toggle = 1}
```

### 2013-06-20 inst/doc/nonparametric-bayes.tex
writing writing

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{Profit <- sims_data[, }\KeywordTok{sum}\NormalTok{(profit), by=}\KeywordTok{c}\NormalTok{(}\StringTok{"reps"}\NormalTok{, }\StringTok{"method"}\NormalTok{)]}
\NormalTok{tmp <- }\KeywordTok{dcast}\NormalTok{(Profit, reps ~ method)}
\NormalTok{tmp <- tmp / tmp[,}\StringTok{"True"}\NormalTok{]}
\NormalTok{tmp <- }\KeywordTok{melt}\NormalTok{(tmp[}\DecValTok{2}\NormalTok{:}\KeywordTok{dim}\NormalTok{(tmp)[}\DecValTok{2}\NormalTok{]])}
\NormalTok{actual_over_optimal <-}\KeywordTok{subset}\NormalTok{(tmp, variable != }\StringTok{"True"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{plot_gp <- }\KeywordTok{ggplot}\NormalTok{(tgp_dat) + }\KeywordTok{geom_ribbon}\NormalTok{(}\KeywordTok{aes}\NormalTok{(x,y,}\DataTypeTok{ymin=}\NormalTok{ymin,}\DataTypeTok{ymax=}\NormalTok{ymax), }\DataTypeTok{fill=}\StringTok{"gray80"}\NormalTok{) +}
    \KeywordTok{geom_line}\NormalTok{(}\DataTypeTok{data=}\NormalTok{models, }\KeywordTok{aes}\NormalTok{(x, value, }\DataTypeTok{col=}\NormalTok{method), }\DataTypeTok{lwd=}\DecValTok{1}\NormalTok{, }\DataTypeTok{alpha=}\FloatTok{0.8}\NormalTok{) + }
    \KeywordTok{geom_point}\NormalTok{(}\DataTypeTok{data=}\NormalTok{obs, }\KeywordTok{aes}\NormalTok{(x,y), }\DataTypeTok{alpha=}\FloatTok{0.8}\NormalTok{) + }
    \KeywordTok{xlab}\NormalTok{(}\KeywordTok{expression}\NormalTok{(X[t])) + }\KeywordTok{ylab}\NormalTok{(}\KeywordTok{expression}\NormalTok{(X[t}\DecValTok{+1}\NormalTok{])) +}
    \KeywordTok{scale_colour_manual}\NormalTok{(}\DataTypeTok{values=}\NormalTok{cbPalette) }
\KeywordTok{print}\NormalTok{(plot_gp)}
\end{Highlighting}
\end{Shaded}

\begin{figure}[htbp]
\centering
\includegraphics{figure/Figure1.pdf}
\caption{Graph of the inferred Gaussian process compared to the true
process and maximum-likelihood estimated process. Graph shows the
expected value for the function $f$ under each model. Two standard
deviations from the estimated Gaussian process covariance with (light
grey) and without (darker grey) measurement error are also shown. The
training data is also shown as black points. (The GP is conditioned on
0,0, shown as a pseudo-data point).}
\end{figure}

@@

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{ggplot}\NormalTok{(actual_over_optimal, }\KeywordTok{aes}\NormalTok{(value)) + }\KeywordTok{geom_histogram}\NormalTok{() + }
  \KeywordTok{facet_wrap}\NormalTok{(~variable, }\DataTypeTok{scales =} \StringTok{"free_y"}\NormalTok{) + }\KeywordTok{guides}\NormalTok{(}\DataTypeTok{legend.position =} \StringTok{"none"}\NormalTok{) + }\KeywordTok{xlab}\NormalTok{(}\StringTok{"Total profit by replicate"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# ggplot(actual_over_optimal, aes(value)) + geom_histogram(aes(fill=variable), binwidth=0.1) + }
\CommentTok{#  xlab("Total profit by replicate")+ scale_fill_manual(values=colorkey)}
\KeywordTok{ggplot}\NormalTok{(actual_over_optimal, }\KeywordTok{aes}\NormalTok{(value, }\DataTypeTok{fill=}\NormalTok{variable, }\DataTypeTok{color=}\NormalTok{variable)) + }
  \KeywordTok{stat_density}\NormalTok{(}\DataTypeTok{position=}\StringTok{"stack"}\NormalTok{, }\DataTypeTok{adjust=}\DecValTok{10}\NormalTok{, }\DataTypeTok{alpha=}\NormalTok{.}\DecValTok{8}\NormalTok{) + }
  \KeywordTok{xlab}\NormalTok{(}\StringTok{"Total profit by replicate"}\NormalTok{)+ }\KeywordTok{scale_fill_manual}\NormalTok{(}\DataTypeTok{values=}\NormalTok{colorkey)+ }\KeywordTok{scale_color_manual}\NormalTok{(}\DataTypeTok{values=}\NormalTok{colorkey)}
\end{Highlighting}
\end{Shaded}

@@

>>>>

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{options}\NormalTok{(}\DataTypeTok{xtable.print.comment =} \OtherTok{FALSE}\NormalTok{)}
\KeywordTok{options}\NormalTok{(}\DataTypeTok{xtable.type =} \StringTok{"latex"}\NormalTok{, }\DataTypeTok{table.placement =} \StringTok{"H"}\NormalTok{)}
\NormalTok{opts_knit$}\KeywordTok{set}\NormalTok{(}\DataTypeTok{progress =} \OtherTok{TRUE}\NormalTok{, }\DataTypeTok{verbose =} \OtherTok{TRUE}\NormalTok{)}
\NormalTok{opts_chunk$}\KeywordTok{set}\NormalTok{(}\DataTypeTok{dev =} \KeywordTok{c}\NormalTok{(}\StringTok{"pdf"}\NormalTok{, }\StringTok{"png"}\NormalTok{), }\DataTypeTok{fig.width =} \FloatTok{5.5}\NormalTok{, }\DataTypeTok{fig.height =} \DecValTok{4}\NormalTok{, }\DataTypeTok{cache.path =} \StringTok{"cache/nonparametric-bayes/"}\NormalTok{, }
    \DataTypeTok{cache =} \OtherTok{TRUE}\NormalTok{, }\DataTypeTok{external =} \OtherTok{TRUE}\NormalTok{)}
\NormalTok{opts_chunk$}\KeywordTok{set}\NormalTok{(}\DataTypeTok{warning =} \OtherTok{FALSE}\NormalTok{, }\DataTypeTok{message =} \OtherTok{FALSE}\NormalTok{, }\DataTypeTok{comment =} \OtherTok{NA}\NormalTok{, }\DataTypeTok{tidy =} \OtherTok{FALSE}\NormalTok{)}
\NormalTok{toggle = }\StringTok{"markup"}
\KeywordTok{theme_set}\NormalTok{(}\KeywordTok{theme_bw}\NormalTok{(}\DataTypeTok{base_size =} \DecValTok{12}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{Profit <- sims_data[, }\KeywordTok{sum}\NormalTok{(profit), by=}\KeywordTok{c}\NormalTok{(}\StringTok{"reps"}\NormalTok{, }\StringTok{"method"}\NormalTok{)]}
\NormalTok{tmp <- }\KeywordTok{dcast}\NormalTok{(Profit, reps ~ method)}
\CommentTok{#tmp$Allen <- tmp[,"Allen"] + rnorm(dim(tmp)[1], 0, 1) # jitter for plotting}
\NormalTok{tmp <- tmp / tmp[,}\StringTok{"True"}\NormalTok{]}
\NormalTok{tmp <- }\KeywordTok{melt}\NormalTok{(tmp[}\DecValTok{2}\NormalTok{:}\KeywordTok{dim}\NormalTok{(tmp)[}\DecValTok{2}\NormalTok{]])}
\NormalTok{actual_over_optimal <-}\KeywordTok{subset}\NormalTok{(tmp, variable != }\StringTok{"True"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{require}\NormalTok{(MASS)}
\NormalTok{step_ahead <- function(x, f, p)\{}
  \NormalTok{h = }\DecValTok{0}
  \NormalTok{x_predict <- }\KeywordTok{sapply}\NormalTok{(x, f, h, p)}
  \NormalTok{n <- }\KeywordTok{length}\NormalTok{(x_predict) - }\DecValTok{1}
  \NormalTok{y <- }\KeywordTok{c}\NormalTok{(x[}\DecValTok{1}\NormalTok{], x_predict[}\DecValTok{1}\NormalTok{:n])}
  \NormalTok{y}
\NormalTok{\}}
\NormalTok{step_ahead_posteriors <- function(x)\{}
\NormalTok{gp_f_at_obs <- }\KeywordTok{gp_predict}\NormalTok{(gp, x, }\DataTypeTok{burnin=}\FloatTok{1e4}\NormalTok{, }\DataTypeTok{thin=}\DecValTok{300}\NormalTok{)}
\NormalTok{df_post <- }\KeywordTok{melt}\NormalTok{(}\KeywordTok{lapply}\NormalTok{(}\KeywordTok{sample}\NormalTok{(}\DecValTok{100}\NormalTok{), }
  \NormalTok{function(i)\{}
    \KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{time =} \DecValTok{1}\NormalTok{:}\KeywordTok{length}\NormalTok{(x), }\DataTypeTok{stock =} \NormalTok{x, }
                \DataTypeTok{GP =} \KeywordTok{mvrnorm}\NormalTok{(}\DecValTok{1}\NormalTok{, gp_f_at_obs$Ef_posterior[,i], gp_f_at_obs$Cf_posterior[[i]]),}
                \DataTypeTok{True =} \KeywordTok{step_ahead}\NormalTok{(x,f,p),  }
                \DataTypeTok{MLE =} \KeywordTok{step_ahead}\NormalTok{(x,f,est$p), }
                \DataTypeTok{Allen =} \KeywordTok{step_ahead}\NormalTok{(x, allen_f, pardist[i,]), }
                \DataTypeTok{Ricker =} \KeywordTok{step_ahead}\NormalTok{(x, ricker_f, ricker_pardist[i,]), }
                \DataTypeTok{Myers =} \KeywordTok{step_ahead}\NormalTok{(x, myers_f, myers_pardist[i,]))}
  \NormalTok{\}), }\DataTypeTok{id=}\KeywordTok{c}\NormalTok{(}\StringTok{"time"}\NormalTok{, }\StringTok{"stock"}\NormalTok{))}
\NormalTok{\}}

\NormalTok{df_post <- }\KeywordTok{step_ahead_posteriors}\NormalTok{(x)}

\KeywordTok{ggplot}\NormalTok{(df_post) + }\KeywordTok{geom_point}\NormalTok{(}\KeywordTok{aes}\NormalTok{(time, stock)) + }
  \KeywordTok{geom_line}\NormalTok{(}\KeywordTok{aes}\NormalTok{(time, value, }\DataTypeTok{col=}\NormalTok{variable, }\DataTypeTok{group=}\KeywordTok{interaction}\NormalTok{(L1,variable)), }\DataTypeTok{alpha=}\NormalTok{.}\DecValTok{1}\NormalTok{) + }
  \KeywordTok{scale_colour_manual}\NormalTok{(}\DataTypeTok{values=}\NormalTok{colorkey, }\DataTypeTok{guide =} \KeywordTok{guide_legend}\NormalTok{(}\DataTypeTok{override.aes =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{alpha =} \DecValTok{1}\NormalTok{))) }
\end{Highlighting}
\end{Shaded}

\begin{figure}[htbp]
\centering
\includegraphics{figure/Figureb.pdf}
\caption{plot of chunk Figureb}
\end{figure}

@@

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{ggplot}\NormalTok{(actual_over_optimal, }\KeywordTok{aes}\NormalTok{(value)) + }\KeywordTok{geom_histogram}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{fill=}\NormalTok{variable)) + }
  \KeywordTok{facet_wrap}\NormalTok{(~variable, }\DataTypeTok{scales =} \StringTok{"free_y"}\NormalTok{)  + }\KeywordTok{guides}\NormalTok{(}\DataTypeTok{legend.position =} \StringTok{"none"}\NormalTok{) +}
  \KeywordTok{xlab}\NormalTok{(}\StringTok{"Total profit by replicate"}\NormalTok{) + }\KeywordTok{scale_fill_manual}\NormalTok{(}\DataTypeTok{values=}\NormalTok{colorkey)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]

\KeywordTok{ggplot}\NormalTok{(actual_over_optimal, }\KeywordTok{aes}\NormalTok{(value)) + }\KeywordTok{geom_histogram}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{fill=}\NormalTok{variable), }\DataTypeTok{binwidth=}\FloatTok{0.1}\NormalTok{) + }
  \KeywordTok{xlab}\NormalTok{(}\StringTok{"Total profit by replicate"}\NormalTok{)+ }\KeywordTok{scale_fill_manual}\NormalTok{(}\DataTypeTok{values=}\NormalTok{colorkey)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]

\KeywordTok{ggplot}\NormalTok{(actual_over_optimal, }\KeywordTok{aes}\NormalTok{(value, }\DataTypeTok{fill=}\NormalTok{variable, }\DataTypeTok{color=}\NormalTok{variable)) + }
  \KeywordTok{stat_density}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{y=}\NormalTok{..density..), }\DataTypeTok{position=}\StringTok{"stack"}\NormalTok{, }\DataTypeTok{adjust=}\DecValTok{3}\NormalTok{, }\DataTypeTok{alpha=}\NormalTok{.}\DecValTok{9}\NormalTok{) + }
  \KeywordTok{xlab}\NormalTok{(}\StringTok{"Total profit by replicate"}\NormalTok{)+ }\KeywordTok{scale_fill_manual}\NormalTok{(}\DataTypeTok{values=}\NormalTok{colorkey)+ }\KeywordTok{scale_color_manual}\NormalTok{(}\DataTypeTok{values=}\NormalTok{colorkey)}
\end{Highlighting}
\end{Shaded}

\begin{figure}[htbp]
\centering
\includegraphics{figure/Figure43.pdf}
\caption{Histograms of the realized net present value of the fishery
over a range of simulated data and resulting parameter estimates. For
each data set, the three models are estimated as described above. Values
plotted are the averages of a given policy over 100 replicate
simulations. Details and code provided in the supplement.}
\end{figure}
